Confronting preferential sampling when analyzing population distributions: diagnosis and model-based triage \footnotemark[2]

### 2017-04-24 manuscript/pref_sampling_revision2.tex
This commit corresponds to code and manuscript status when the manuscript was accepted at MEE. It represents the first commit where cross validation was used in the sim example.

\def\volumeyear{2016}
\def\VOC{VO}

\begin{document}
\runningheads{P.~B.~Conn, J.~T.~Thorson, \& D.~S.~Johnson}{Preferential sampling in species distribution models}

@@

\title{Confronting preferential sampling in wildlife surveys: diagnosis and model-based triage \footnotemark[2]}

@@

\begin{abstract}
 %\noindent {\emph{Summary}}
\small
\begin{enumerate}
\item  Wildlife surveys are often used to estimate the density, abundance, or distribution of animal populations. Recently, model-based approaches to analyzing survey data have become popular because one can more readily accommodate departures from pre-planned survey routes and construct more detailed maps than one can with design-based procedures.

\item Species distribution models fitted to wildlife survey data often make the implicit assumption that locations chosen for sampling and animal abundance at those locations are conditionally independent given modeled covariates.  However, this assumption may be violated when survey effort is non-randomized, leading to preferential sampling.

\item We develop a hierarchical statistical modeling framework for detecting and alleviating the biasing effects of preferential sampling in species distribution models fitted to count data.  The approach works by jointly modeling wildlife state variables and the locations selected for sampling and specifying a dependent correlation structure between the two models.

@@

\item When wildlife surveys are conducted without a well-defined sampling frame, ecologists should recognize the potentially biasing effects of preferential sampling when estimating animal density or occurrence. Joint models, such as those described in this paper, can be used to test and correct for such biases. However, such models can be unstable; ultimately the best way to avoid preferential sampling bias is to incorporate design-based principles such as randomization and/or systematic sampling into survey design.
\end{enumerate}

@@

\keywords{count data, preferential sampling, spatial autocorrelation, species distribution model}

@@

Surveys of unmarked animal populations are often used to estimate abundance and occurrence of animal populations and to predict species distributions, enterprises central to conservation, ecology, and management. For studies of abundance, researchers historically relied on design-based statistical inference \citep[e.g.][]{Cochran1977}, which requires adoption of a pre-defined sampling frame (e.g. using systematic random sampling, stratified random sampling, or some variant thereof).  Designing animal surveys is relatively straightforward in such applications, and unbiased point and variance estimators are available.  Recently, however, there has been a surge in research describing model-based procedures for estimating abundance, density, and occupancy from surveys of unmarked animals, including N-mixture and Dail-Madsen models for repeated point counts \citep{Royle2004a,DailMadsen2011}, occupancy models for presence-absence surveys \citep{MacKenzie2002,JohnsonEtAl2013}, and various model-based formulations for distance-sampling data \citep{HedleyBuckland2004,JohnsonEtAl2010,MillerEtAl2013}.  In such applications, it is common to use habitat or environmental covariates together with spatial effects (e.g. via trend surfaces or spatial random effects) to predict density or distributions across the landscape.  We shall refer to the amalgam of model-based approaches for making spatially explicit inference about animal populations as ``species distribution models" \citep[SDMs; {\it sensu}][]{ElithLeathwick2009}, even though this term is more often used to refer to animal occurrence than it is to density or abundance.

One of the main advantages of using SDMs is that one is less beholden to predetermined sampling frames and can potentially use data from non-randomized designs or opportunistic sampling to make inferences about animal populations \citep{JohnsonEtAl2010}. Several authors have shown that non-randomized designs can have poorer predictive ability than randomized designs when used to estimate species distributions \citep[e.g.][]{EdwardsEtAl2006,AlbertEtAl2010}, but that the optimal design may depend on the question of interest.  For instance, the optimal design may differ depending on whether the inferential goal is to maximize predictive ability or to estimate the relationship between species presence and important habitat covariates \citep{AlbertEtAl2010}.

Less precise predictions may be a sacrifice one is willing to make if it is the only cost incurred when using data at non-random locations to make inferences about species distributions.  After all, such data are often all that are available for certain taxa and locations (e.g. from citizen science programs), and are certainly less expensive to acquire than data from randomized, scientific surveys.  However, recent research suggests that one may be opening oneself up to bias as well, which is potentially much more concerning than loss of precision. In a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.  In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance. For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.  Bias might also arise if survey effort is higher near bases of operations or features such as roads, particularly if animal abundance is higher (or lower) near these locations than elsewhere in the landscape.  In our view, this phenomenon, termed ``preferential sampling," is potentially much more problematic, because estimates of animal abundance and occurrence from SDMs could be severely overstated.

In this article, we explore potential for bias in SDMs resulting from preferential sampling (hereafter, PS) and describe several model-based approaches for detecting and correcting for such biases.  We start by describing a common currency for notation and basic model structures considered in this paper.  Second, we review PS bias in a mathematical light, and describe prior approaches to coping with its effects. Third, we introduce a novel generalization of previously proposed PS models, allowing the investigator to jointly model animal encounter data and the locations chosen for sampling, including possible dependence structure between these two types of observations.  Fourth, we conduct a simulation study to examine the performance of traditional SDMs and our newly developed PS model when data are gathered preferentially.  Finally, we demonstrate our modeling approach by analyzing aerial survey counts of bearded seals (\textit{Erignathus barbatus}) in the Bering Sea.

@@

We focus here exclusively on discrete space (areal) models for animal encounter data as these seem to be the dominant form used in design and analysis of animal population surveys, although we note that PS is likely to affect analyses similarly regardless of the choice of spatial domain.  We suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$). Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if location $i$ is sampled (i.e. if $U_i \in \mathcal{S}$), and is 0 otherwise.  To formulate a ``traditional" SDM, one could then write animal abundance or occurrence as a stochastic realization of a probability mass function $f()$:
\begin{eqnarray}
  \label{eq:process}
  [Z_i] & = & f(g^{-1}(\mu_i)),
\end{eqnarray}
where we use the bracket notation ($[\cdot]$) to denote a probability distribution (density) function.  In this example, $Z_i$ denotes the state variable of interest (e.g. occupancy or abundance), $g()$ is a link function (e.g. probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value which can itself be written as a function of habitat covariates, regression coefficients, and spatially autocorrelated random effects.  Spatially autocorrelated random effects are often included to allow animal density to vary smoothly in space and to account for residual patchiness not explained by predictive covariates \citep{Legendre1993,LichsteinEtAl2002}. In applications described in this paper, we write the intensity as
\begin{eqnarray}
\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i,
 \label{eq:mu}
\end{eqnarray}
where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, and $\boldsymbol{\delta}=\{ \delta_1,\delta_2,\hdots,\delta_S \}$ are spatially autocorrelated random effects.  For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\delta_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g. conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{ReichEtAl2006,Hughes2013}.

The model for $Z_i$ describes variation in the process of interest and is often described as the ``process'' model.  However, it is usually impossible to conduct a complete census of all animals present in a surveyed location, so it is customary to include an observation model describing incomplete detection.  For occupancy studies, the response variable is $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}:
\begin{eqnarray}
  [Y_i|Z_i] & = & \text{Bernoulli}(Z_i p_i).
\end{eqnarray}
 Here, the detection probability $p_i$ is possibly a function of survey and observer specific covariates (e.g. visibility, observer experience). Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$.   For count surveys, a possible model is
\begin{eqnarray}
  \label{eq:obs_binomial}
  [Y_i|Z_i] & = & \text{Binomial}(Z_i, A_i p_i),
\end{eqnarray}
where the $Y_i$ now represents the count of animals obtained while surveying unit $i$, $A_i$ denotes the proportion of sample unit $i$ that is surveyed, and $p_i$ gives detection probability.  Additional information will often be needed to estimate $p_i$ in this context, such as data from double observers, distance observations, or double sampling \citep[see e.g.][]{BucklandEtAl2001,Royle2004,Borchers2006,ConnEtAl2014}.

For the remainder of this treatment, we use bold symbols to denote vector-valued quantities or matrices.  We also use ``$|$" to specify that a distribution is conditional; for instance, if $[{\bf Y}]$ denotes the marginal probability mass function for ${\bf Y}$, then  $[{\bf Y}|{\bf Z}]$ represents the conditional distribution of ${\bf Y}$ given ${\bf Z}$.  We use $\bfmu$ and $\bfnu$ to indicate the link-scale process of interest and the logit of the probability of sampling, respectively.  For animal abundance, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects.  We use the notation $Z_i$ when describing the process model in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.

@@

Design-based estimation requires that surveys adhere to a pre-planned survey design selected probabilistically from an underlying sampling frame \citep{Cochran1977}.  Although randomization can increase legitimacy, model-based estimation is not formally subject to this requirement.  For instance, investigators can reallocate sampling effort if weather or logistics preclude surveying in a desired location and still generate model-based estimates of abundance or occupancy.  This can be a crucial advantage in surveys covering large areas with frequent inclement weather.  It also opens the door for using presence-only, citizen science, or opportunistically collected data for estimation.

However, the manner in which effort is allocated can potentially have profound influence on SDM estimator performance.   Preferential sampling (PS) arises when the locations being sampled and the process of interest (e.g. density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.  For instance, PS can occur when the investigator tends to sample more in places where a species has already been observed, or is known to be abundant. \citet{DiggleEtAl2010} showed that this type of PS can lead to bias when this extra information is not included in models for the state variable of interest.  Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$, where \textbf{R} is an indicator vector whose elements $R_i$ are 1.0 if sampling unit $i$ is sampled and are zero otherwise (see e.g. fig. \ref{fig:PS_example}).  We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.  For instance, \citet{MerckxEtAl2011} use the term ``preferential sampling" to refer to the process of visiting some sites more often than others, while \citet{ManceurKuhn2014} define it as occurring when the locations selected for sampling are a function of an environmental covariate.  Neither of these latter conditions are problematic outside of the specialized field of presence-only modelling.

@@

As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS (e.g. fig. \ref{fig:PS_example}).  However, in many ecological applications, predictive covariates explain only a small portion of variation present in the data.  If the locations selected for sampling are a function of some unmodelled factor related to abundance (intentionally or unintentionally), bias may still occur.  Despite the clear potential for bias in SDMs, there are few examples where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.  One exception is
\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs.  However, they did not attempt to account for PS in their models.

@@

Several authors have attempted model-based corrections for PS in the statistical literature. For Gaussian models on a continuous spatial domain, \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} jointly modeled the process of interest and the locations chosen for sampling.  In particular, they expressed sampled locations as an inhomogeneous Poisson point process where the underlying log-scale intensity at a given location depended linearly on the process of interest at that location.  For instance, writing observations of the spatial process (random field) at a location $i$ as
\begin{eqnarray}
   Z_i & = & \mu_i + \epsilon_i,
   \label{eq:Z_cont}
\end{eqnarray}
the log-scale relative intensity of sampling locations at $i$ ($\nu_i$) could be written as
\begin{eqnarray}
   \nu_i & \propto & \xi_i + b \mu_i,
   \label{eq:p_cont}
\end{eqnarray}
where $\mu_i$ and $\xi_i$ can both be written as a function of predictive covariates.
Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the spatial process (e.g. animal density) is high, and $b<0$ implies greater sampling where the spatial process is low.  Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that ``. . . accounting for informative sampling is only necessary when there
is an association between the spatial surface of interest and the sampling density that cannot be
explained by the shared spatial covariates" (see e.g. fig. \ref{fig:PS_example}).    \citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, log sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in eqn. (\ref{eq:Z_cont}), finding that this approach helped reduce bias associated with PS, but did not perform as well as the full joint model.

@@

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurrence) and the locations chosen for sampling. For the abundance process model, we start with eqn. (\ref{eq:process}) as a general formulation for non-Gaussian data, writing the link-scale expectation as in eqn. (\ref{eq:mu}).
Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is selected for sampling, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
\begin{eqnarray}
 \label{eq:R}
  [R_i] & = & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \boldsymbol{\nu} & = & \beta_0^* + {\bf x}^* \boldsymbol{\beta}^*+\boldsymbol{\eta}+{\bf B} \boldsymbol{\delta}.
\label{eq:nu}
\end{eqnarray}
  As with the process model, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$) and spatially autocorrelated random effects ($\boldsymbol{\eta}$ and $\boldsymbol{\delta}$).  The predictive covariates ${\bf x}_i$ from eqn. (\ref{eq:mu}) and ${\bf x}_i^*$ from eqn. (\ref{eq:nu}) may or may not be the same.  Note also that the  spatially autocorrelated random effects $\boldsymbol{\delta}$ are included in both eqns. (\ref{eq:mu}) and (\ref{eq:nu}), allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between sampling intensity and the process of interest.  The spatially autocorrelated random effects $\bfeta$ are assumed independent of the $\bfdelta$.

>>>>

\def\volumeyear{2017}
\def\VOC{VO}

\begin{document}
\runningheads{P.~B.~Conn, J.~T.~Thorson, \& D.~S.~Johnson}{Preferential sampling in spatial distribution models}

@@

\title{Confronting preferential sampling when analyzing population distributions: diagnosis and model-based triage \footnotemark[2]}

@@

\begin{abstract}
 %\noindent {\emph{Summary}}
\small
\begin{enumerate}
\item  Population surveys are often used to estimate the density, abundance, or distribution of natural populations. Recently, model-based approaches to analyzing survey data have become popular because one can more readily accommodate departures from pre-planned survey routes and construct more detailed maps than one can with design-based procedures.

\item Spatial models for population distributions often make the implicit assumption that locations chosen for sampling and animal abundance at those locations are conditionally independent given modeled covariates.  However, this assumption may be violated when survey effort is non-randomized, leading to preferential sampling.

\item We develop a hierarchical statistical modeling framework for detecting and alleviating the biasing effects of preferential sampling in spatial distribution models fitted to count data.  The approach works by specifying a joint model for population density and the locations selected for sampling, and specifying a dependent correlation structure between the two processes.

@@

\item When population surveys break from traditional scientific survey design principles, ecologists should recognize the potentially biasing effects of preferential sampling when estimating population density or occurrence. Joint models, such as those described in this paper, can be used to test and correct for such biases. However, such models can be unstable; ultimately the best way to avoid preferential sampling bias is to incorporate design-based principles such as randomization and/or systematic sampling into survey design.
\end{enumerate}

@@

\keywords{count data, population survey, preferential sampling, spatial autocorrelation, spatial distribution model, species distribution model}

@@

Ecologists often fit statistical models to data from surveys of natural populations (e.g., plants, animals) to explain variation in species distributions and to predict presence or abundance in unsampled areas. Such models may serve as the basis for testing ecological theory \citep{HilbornMangel1997,Austin2007}, for directed conservation and management \citep{WilliamsEtAl2002,NicholsWilliams2006}, or for monitoring biological diversity \citep{PereiraCooper2006,YoccozEtAl2001}.  Depending on the application, data quality can vary widely, ranging from scientific survey data to opportunistic records of species presence reported by the public.

Where possible, researchers often use sound survey design principles (e.g. randomization) to ensure unbiased inference when implementing scientific surveys.  For instance, under protocols such as simple, systematic, or stratified random sampling, design-based estimators of average density or presence-absence over large survey regions are guaranteed to be unbiased and avoid making assumptions about the factors influencing density or distribution \citep{Cochran1977,BucklandEtAl2000}.  However, spatial resolution is typically poor for such estimators, limiting the breadth of ecological and management questions that can be addressed. Further, for logistical reasons, many ecological surveys simply cannot afford to be this rigorous in their design.  For example, even the North American Breeding Bird Survey uses roads and other non-random features that necessitate use of alternate estimation procedures \citep{SauerLink2011}.

Over the last decade, there has thus been a huge surge in research describing model-based procedures for estimating abundance, density, or presence-absence from population survey data.  In contrast to design-based procedures, model-based procedures make explicit assumptions about how populations vary over space, both in terms of explanatory variables and often in terms of distribution.  For instance, in such applications, it is common to use habitat or environmental covariates together with spatial effects (e.g. via trend surfaces or spatial random effects) to predict density or distributions across the landscape. This approach can be useful for fine-scale management, identifying range shifts, and comparing the fit (or lack thereof) of alternative models that express different ecological hypotheses \citep{Austin2007}.  We shall refer to the amalgam of model-based approaches for making spatially explicit inference about natural populations as ``spatial models for population distributions" (SMPDs).  Ultimately, the quality of inference from SMPDs depends on several factors, including (i) how well the fitted model approximates reality, and (ii) the quality of the data \citep{BucklandEtAl2000}.

The radiation of model-based approaches has proceeded at different rates and pathways in different ecological modelling subfields, which are often linked to the data being modeled.  For instance, species distribution models \citep[SDMs; ][]{ElithLeathwick2009} are often used to predict species presence as a function of habitat variables and encompass both statistical (e.g. generalized linear or generalized additive models) and machine learning algorithms. The term SDM also connotes analysis of presence-only data (as available from museum collection or citizen science programs for instance).  In the field of population assessment, statistical models are almost exclusively used, often with observation models tailored to the sampling approach (e.g. presence-absence surveys, distance sampling, quadrat samples, point counts).  In this paper, our perspective is one shaped more heavily by the population assessment literature, though we expect our general approach to be applicable to a wider variety of SMPDs.

One of the main advantages of using SMPDs is that one can potentially use data from non-randomized designs or opportunistic sampling to make inferences about animal populations \citep{JohnsonEtAl2010}. Several authors have shown that non-randomized designs can have poorer predictive ability than randomized designs when used to estimate species distributions \citep[e.g.][]{EdwardsEtAl2006,AlbertEtAl2010}, but that the optimal design may depend on the question of interest.  For instance, the optimal design may differ depending on whether the inferential goal is to maximize predictive ability or to estimate the relationship between species presence and important habitat covariates \citep{AlbertEtAl2010}.

Less precise predictions may be a sacrifice one is willing to make if it is the only cost incurred when using data at non-random locations to make inferences about species distributions.  After all, such data are often all that are available for certain taxa and locations (e.g. from citizen science programs), and are certainly less expensive to acquire than data from randomized, scientific surveys.  However, recent research suggests that one may be opening oneself up to bias as well, which is potentially much more concerning than loss of precision. In a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.  In the context of SMPDs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance. For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.  Bias might also arise if survey effort is higher near bases of operations or features such as roads, particularly if animal abundance is higher (or lower) near these locations than elsewhere in the landscape.  In our view, this phenomenon, termed ``preferential sampling," is potentially much more problematic, because estimates of animal abundance and occurrence from SMPDs could be severely overstated.

In this article, we explore potential for bias in SMPDs resulting from preferential sampling (hereafter, PS) and describe several model-based approaches for detecting and correcting for such biases.  We start by describing a common currency for notation and basic model structures considered in this paper.  Second, we review PS bias in a mathematical light, and describe prior approaches to coping with its effects. Third, we introduce a novel generalization of previously proposed PS models, allowing the investigator to jointly model animal encounter data and the locations chosen for sampling, including possible dependence structure between these two types of observations.  Fourth, we conduct a simulation study to examine the performance of traditional SDMs and our newly developed PS model when data are gathered preferentially.  Finally, we demonstrate our modeling approach by analyzing aerial survey counts of bearded seals (\textit{Erignathus barbatus}) in the Bering Sea.

@@

We focus here exclusively on discrete space (areal) models for population survey data, although we note that PS is likely to affect analyses similarly regardless of the choice of spatial domain.  We suppose that the investigator intending to fit a SMPD to survey data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are sampled (call the set of sampled locations $\mathcal{S}$). Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if location $i$ is sampled (i.e. if $U_i \in \mathcal{S}$), and is 0 otherwise.  To formulate a ``traditional" SMPD, one could then write abundance or occurrence as a stochastic realization of a probability mass function $f()$:
\begin{eqnarray}
  \label{eq:process}
  [Z_i] & = & f(g^{-1}(\mu_i)),
\end{eqnarray}
where we use the bracket notation ($[\cdot]$) to denote a probability distribution (density) function.  In this example, $Z_i$ denotes the state variable of interest (e.g. occupancy or abundance), $g()$ is a link function (e.g. probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value which can itself be written as a function of habitat covariates, regression coefficients, and spatially autocorrelated random effects.  Spatially autocorrelated random effects are often included to allow populations to vary smoothly in space and to account for residual patchiness not explained by predictive covariates \citep{Legendre1993,LichsteinEtAl2002}. In applications described in this paper, we write the intensity as
\begin{eqnarray}
\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i,
 \label{eq:mu}
\end{eqnarray}
where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, and $\boldsymbol{\delta}=\{ \delta_1,\delta_2,\hdots,\delta_S \}$ are spatially autocorrelated random effects.  For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\delta_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g. conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{ReichEtAl2006,Hughes2013}.

The model for $Z_i$ describes variation in the process of interest and is often described as the ``process'' model.  However, it is usually impossible to conduct a complete census of all individuals present in a surveyed location, so it is customary to include an observation model describing incomplete detection.  For occupancy (presence/absence) studies, the response variable is $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}:
\begin{eqnarray}
  [Y_i|Z_i] & = & \text{Bernoulli}(Z_i p_i).
\end{eqnarray}
 Here, the detection probability $p_i$ is possibly a function of survey and observer specific covariates (e.g. visibility, observer experience). Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$ \citep{MacKenzie2002}.   For count surveys, a possible model is
\begin{eqnarray}
  \label{eq:obs_binomial}
  [Y_i|Z_i] & = & \text{Binomial}(Z_i, A_i p_i),
\end{eqnarray}
where the $Y_i$ now represents the count of animals obtained while surveying unit $i$, $A_i$ denotes the proportion of sample unit $i$ that is surveyed, and $p_i$ gives detection probability.  Additional information will often be needed to estimate $p_i$ in this context, such as data from double observers, distance observations, or double sampling \citep[see e.g.][]{BucklandEtAl2001,Royle2004,Borchers2006,ConnEtAl2014}.

For the remainder of this treatment, we use bold symbols to denote vector-valued quantities or matrices.  We also use ``$|$" to specify that a distribution is conditional; for instance, if $[{\bf Y}]$ denotes the marginal probability mass function for ${\bf Y}$, then  $[{\bf Y}|{\bf Z}]$ represents the conditional distribution of ${\bf Y}$ given ${\bf Z}$.  We use $\bfmu$ and $\bfnu$ to indicate the link-scale process of interest and the logit of the probability of sampling, respectively.  For population abundance, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects.  We use the notation $Z_i$ when describing the process model in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.

@@

Design-based estimation requires that surveys adhere to a pre-planned survey design selected probabilistically from an underlying sampling frame to ensure unbiasedness \citep{Cochran1977}.  Although randomization can reduce potential for bias, researchers often view SMPDs in a different light, where the data gathering procedure is not important as long as the models being fitted are formulated well and produce reasonable predictions (see \textit{Discussion} for more on this topic).  Under this view, investigators can reallocate effort if weather or logistics preclude surveying in a desired location and still generate model-based estimates of abundance or occupancy.  This can be a crucial advantage in surveys covering large areas with frequent inclement weather.  It also opens the door for using presence-only, citizen science, or opportunistically collected data for estimation.

However, the manner in which effort is allocated can potentially have profound influence on SMPD estimator performance.   Preferential sampling (PS) arises when the locations being sampled and the process of interest (e.g. density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.  For instance, PS can occur when the investigator tends to sample more in places where a species has already been observed, or is known to be abundant. \citet{DiggleEtAl2010} showed that this type of PS can lead to bias when this extra information is not included in models for the state variable of interest.  Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$, where \textbf{R} is an indicator vector whose elements $R_i$ are 1.0 if sampling unit $i$ is sampled and are zero otherwise (see e.g. fig. \ref{fig:PS_example}).  We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.  For instance, \citet{MerckxEtAl2011} use the term ``preferential sampling" to refer to the process of visiting some sites more often than others, while \citet{ManceurKuhn2014} define it as occurring when the locations selected for sampling are a function of an environmental covariate.  Neither of these latter conditions are problematic outside of the specialized field of presence-only modelling.

@@

As in the pollution example, having good explanatory covariates may also reduce bias when fitting SMPDs to population survey data under PS (e.g. fig. \ref{fig:PS_example}).  However, in many ecological applications, predictive covariates explain only a small portion of variation present in the data.  If the locations selected for sampling are a function of some unmodelled factor related to abundance (intentionally or unintentionally), bias may still occur.  Despite the clear potential for bias in SMPDs, there are few examples where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SMPDs.  One exception is
\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SMPDs to data obtained using nonrandomized designs.  However, they did not attempt to account for PS in their models.

@@

Several authors have attempted model-based corrections for PS in the statistical literature. For Gaussian models on a continuous spatial domain, \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} jointly modeled the process of interest and the locations chosen for sampling.  In particular, they expressed sampled locations as an inhomogeneous Poisson point process where the underlying log-scale intensity at a given location depended linearly on the process of interest at that location.  For instance, writing observations of the spatial process (random field) at a location $i$ as
\begin{eqnarray}
   Z_i & = & \mu_i + \epsilon_i,
   \label{eq:Z_cont}
\end{eqnarray}
the log-scale relative intensity of sampling locations at $i$ ($\kappa_i$) could be written as
\begin{eqnarray}
   \kappa_i & \propto & \xi_i + b \mu_i,
   \label{eq:p_cont}
\end{eqnarray}
where $\mu_i$ and $\xi_i$ can both be written as a function of predictive covariates.
Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the spatial process (e.g. population density) is high, and $b<0$ implies greater sampling where the spatial process is low.  Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that accounting for preferential sampling  ``. . . is only necessary when there
is an association between the spatial surface of interest and the sampling density that cannot be
explained by the shared spatial covariates" (see e.g. fig. \ref{fig:PS_example}).    \citet{PatiEtAl2011} also consider a simpler, plug-in based estimator in which log sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in eqn. (\ref{eq:Z_cont}), finding that this approach helped reduce bias associated with PS, but did not perform as well as the full joint model.

@@

Given these limitations, our present task is to generalize PS models to the types of data more typical of SMPDs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (abundance or occurrence) and the locations chosen for sampling. For the abundance process model, we start with eqn. (\ref{eq:process}) as a general formulation for non-Gaussian data, writing the link-scale expectation as in eqn. (\ref{eq:mu}).
Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is sampled, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
\begin{eqnarray}
 \label{eq:R}
  [R_i] & = & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \boldsymbol{\nu} & = & \beta_0^* + {\bf x}^* \boldsymbol{\beta}^*+\boldsymbol{\eta}+{\bf B} \boldsymbol{\delta}.
\label{eq:nu}
\end{eqnarray}
  As with the process model, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$) and spatially autocorrelated random effects ($\boldsymbol{\eta}$ and $\boldsymbol{\delta}$).  The predictive covariates ${\bf x}_i$ from eqn. (\ref{eq:mu}) and ${\bf x}_i^*$ from eqn. (\ref{eq:nu}) may or may not be the same.  Note also that the  spatially autocorrelated random effects $\boldsymbol{\delta}$ are included in both eqns. (\ref{eq:mu}) and (\ref{eq:nu}), allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between sampling intensity and the process of interest.  The spatially autocorrelated random effects $\bfeta$ are assumed independent of the $\bfdelta$.

 The formulation in eqn. (\ref{eq:nu}) is similar (but not identical) to one proposed by \citet{RoyleBerliner1999} for hierarchical multivariate models with spatial dependence.  There are multiple ways of structuring ${\bf B}$ depending on the complexity of spatial dependence desired for the PS process \citep{RoyleBerliner1999}.  For instance, setting ${\bf B} = {\bf 0}_{S \times S}$ (i.e. to an all-zero matrix) corresponds to an absence of spatial dependence (and thus no PS).  Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear PS model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.  Alternatively, we could allow the degree of PS to vary across the landscape.  For instance, one could contemplate a trend surface model for PS by specifying a matrix for ${\bf B}$, with diagonal entries given by
 \begin{eqnarray*}
   \label{eq:ts}
   B_{ii} & = & b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i
 \end{eqnarray*}
 and $B_{ij}=0$ for $i \ne j$.  Here, $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.  To our knowledge, the identifiability of higher order models (such as trend surfaces) has not been investigated.
